---
title: "R course: data management"
author: "Carlos Ruiz"
date: "30 May 2017"
output:
  BiocStyle::html_document:
    number_sections: true
    toc: yes
---

# Introduction 

R was designed to ease performing statistical analyses. Consequently, R comes with a lot of native functions to manage and clean your data. In this document, I will show how to use these functions using some example data.

# Read data

We can read tabular data in R using the function `read.table`. This is the most generic function and it can be tuned to read any format. In addition, there are some other functions designed to read specific formats. For instance, `read.csv` reads csv files and `read.delim` reads tab delimited files: 

```{r, eval = FALSE}
df <- read.csv("example.csv")
df <- read.delim("example.tab")
```

R can also read files created with other programs, such as .dta files from STATA or .xlsx files from Excel. However, these functions are in other packages. We can load STATA files with the function `read.dta` of foreign package:

```{r, eval = FALSE}
library(foreign)
df <- read.dta("example.dta")
```

To load xlsx files, we can use the function `read.xlsx` from package xlsx:

```{r, eval = FALSE}
library(xlsx)
df <- read.xlsx("example.xlsx")
```

## Examine our data

In the course folder, we have included two csv files. These files contain the data that we will use in our example:

```{r}
dades <- read.csv("dades.csv")
muni <- read.csv("data_muni.csv")
```

The data.frame `dades` contains some measurements performed in different Spanish municipalities. We can examine our data using the function `View`:

```{r}
View(dades)
```

The data.frame `muni` contains some information about the municipalities:

```{r}
View(muni)
```


# Subset by variable

A common operation in data cleaning is to select only the registers of a table that fulfill a requirement. One option to do so is the function `subset`. In this example, we will use the data.frame `muni`. Our first step will be select all registers from Euskadi:

```{r}
subset(muni, ccaaName == "Euskadi")
```

The function `subset` has two main arguments. The first is a data.frame and the second is the expression used to filter. In our case, we want all registers whose `ccaaName` is equal to Euskadi. Remark that the column name goes **without quotes** and that to say equal to we write `==`. We can also use expressions such as greater than or smaller than. For instance, we can select registers by population size:

```{r}
subset(muni, Inhabitants > 200000)
subset(muni, Inhabitants < 2000)
```

We can also use build more complex expression using logical operators:

- `&` means AND
- `|` means OR

For instance, we can select registers from Euskadi AND from municipalities with more than 100000 inhabitants:

```{r}
subset(muni, ccaaName == "Euskadi" & Inhabitants > 100000)
```

We can also select municipalities from Euskadi OR smaller than 2000 inhabitants:

```{r}
subset(muni, ccaaName == "Euskadi" | Inhabitants < 2000)
```

# Create new variables

One the most common steps in data cleaning is creating new variables. We can add a new variable to a data.frame using the `$` and `<-` operators. For instance, we can add a new column full of 0s in `muni`: 

```{r, echo = FALSE}
muniori <- muni
```


```{r}
muni$zero <- 0
muni
```


```{r, echo = FALSE}
muni <- muniori
```

## Dummy variables

The function `ifelse` can be used to create dummy variables. Let's create a dummy variable indicating if a municipality is placed in Albacete:

```{r}
muni$Albacete <- ifelse(muni$ProvinciaName == "Albacete", 1, 0)
muni
```

```{r, echo = FALSE}
muni <- muniori
```

The function `ifelse` has three parameters. The first, is a logical expression, the second the result if the expression is TRUE and the third the result if the expression is FALSE. Then, `ifelse` evaluates the expression for each element of the vector. If an element is TRUE, at this position we get the result of the second parameter. If it is FALSE, we get the result of the third parameter. Let's have a look at a simpler example:

```{r}
vec <- c("a", "b", "a", "a", "c")
vec
ifelse(vec == "a", "Nice", "Bad")
```

## From continuous to categorical

As we can see, `ifelse` can return other results than 0s and 1s. If we use this function with a continuous variable, we can easily create a categorical variable with two levels. Let's create a variable indicating if a municipality is a town or a city depending on the number of inhabitants:

```{r}
muni$muniType <- ifelse(muni$Inhabitants > 100000, "City", "Town")
muni
```

We can concatenate `ifelse` statements to get more complex results. For instance, we can classify municipalities in three categories (village, town, city):

```{r}
muni$muniType2 <- ifelse(muni$Inhabitants > 100000, "City", 
                        ifelse(muni$Inhabitants < 5000, "Village", "Town"))
muni
```

Let's take a look at this code more carefully. The first `ifelse` classifies as cities the municipalities with more than 100000 inhabitants. For the others, we apply a second `ifelse`. If these municipalities have less than 5000 inhabitants, they are village. If not, they are towns. 

This way of creating new categorical variables is useful when there are few levels and the limits are clear. However, sometimes we may be interested in creating the categorical variables using percentiles. To do so, we can use the function `cut`. Let's classify towns using terciles of their mean Temperature:

```{r}
muni$Temp <- cut(muni$MeanT, 3)
muni
```

The first parameter of `cut` is a numeric vector and the second the breaks. We can pass a single number with the number of groups to be done (i.e. 3 for terciles, 4 for quartiles...) or the cut points for the groups. In our first use of `cut`, the names of the groups were the intervals of each group. We can change this using the parameter `labels`:
```{r}
muni$Temp2 <- cut(muni$MeanT, 3, c("Cold", "Mild", "Warm"))
muni
```

The groups are the same but the names are clearer. 

## Merge variables

Another common operation is the create a new variable that is a merge of different variables. In our case, we will create the variable `wholeDate` by merging the variables `dd`, `mm` and `yyyy`. We can do this using the function `paste`:

```{r}
dades$wholeData <- paste(dades$dd, dades$mm, dades$yyyy, sep = "/")
dades
```

We pass all the character vectors we want to merge to the function `paste` separated by commas. The parameter `sep` has the character used to separate the different parts. For instance, can change it by a `-`:

```{r}
dades$wholeData <- paste(dades$dd, dades$mm, dades$yyyy, sep = "-")
dades
```

`paste` always returns a character vector. However, in this case, we might be interested in having a vector of class `Date`, which has some other interesting features. To do so, we should create a vector with Year-Month-Day:

```{r}
dades$wholeData <- as.Date(paste(dades$yyyy, dades$mm, dades$dd, sep = "-"))
dades
```

# Get aggregate data

In this last part, we will show how to get aggregate your data by a variable. Let's imagine that we are interested in having some statistics of our municipalities by province. We can do this using the function `aggregate`:

```{r}
aggregate(muni[, 7:8], list(Provincia = muni$ProvinciaName), mean)
```

In this case, we only selected the columns `Inhabitants` and `MeanT` as they are the only continuous variables. The first element is the data.frame. The second, a **list** with the variables used to aggregate. It should be named to have a proper result. Finally, we have the function to apply. In our case, `mean`. We can add more variables to aggregate the data. Let's add the variable `muniType2` with the classification by size:

```{r}
aggregate(muni[, 7:8], list(Provincia = muni$ProvinciaName, 
                            Size = muni$muniType2), mean)
```

# Search duplicates

We can look for duplicate data using the function `duplicated`. This function scans a vector for duplicates and returns a logical vector. If a value is the first time that appears, its value is FALSE. If it has already appeared, its value is TRUE. Let's check it in `dades` data.frame:

```{r}
dades$muncode[c(1:3, 16:18)]
duplicated(dades$muncode[c(1:3, 16:18)])
```

We can count the number of duplicates by applying the function `sum` to the result of `duplicated`: 
```{r}
sum(duplicated(dades$muncode))
sum(duplicated(muni$muncode))
```

While there are 15 registers that have a repeated muncode in `dades`, there is only one register per muncode in `muni`. We can apply `duplicated` to a data.frame. In this case, a register will be duplicated only if all columns are equal: 


```{r}
dades[c(1:3, 16:18), 1:2]
duplicated(dades[c(1:3, 16:18), 1:2])
dades[c(1:3, 16:18), 1:3]
duplicated(dades[c(1:3, 16:18), 1:3])

```

# Merge data.frames

The first step that we will perform will be merging the two data.frames. As a result, we will have in one data.frame the measurements and the information about the towns. 

There are different options to merge data.frames in R. The most general option is to use the function `merge`. By default, this function merges the data.frames using the rownames. However, in our case, we want to merge the data.frames using the `muncode` column. We can change this with the parameter `by`: 

```{r}
all <- merge(dades, muni, by = "muncode", sort = FALSE)
head(all)
```

If the name of the columns used to merge are different, we can set them using `by.x` and `by.y`. We can also choose what to do if there are values that are found in one data.frame or not the other. Finally, we can sort the output using the merging column (parameter `sort`).

The function `cbind` binds two data.frames by columns, i.e. the columns of one data.frame are added to another data.frame. The main requirement is that both data.frames have the same number of rows:

```{r}
dades[1:5, 1:3]
cbind(dades[1:5, 1:3], dades[1:5, 1:3])
```

`rbind` is a function analogous to `cbind` that binds data.frames by rows: 

```{r}
dades[1:3, ]
rbind(dades[1:3, ], dades[1:3, ])
```

# Write output

The main function to write data to a file is `write.table`. We also have the function `write.csv` to generate a csv file. Both functions share most of the arguments, so we will comment `write.table`. Let's write our new `muni` data.frame to a new file:

```{r}
write.table(muni, file = "newMuni.txt")
```

The two main arguments of write.table are the data.frame and the name of the output file. With the default parameters, the new file have row and column names, the characters are quoted and the fields are separated by a space. We can customize all this options. In our case, we will remove the quotes and the row names and we will change the field separator by a tab:

```{r}
write.table(muni, file = "newMuni2.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

`sep` indicates the character used to separate fields, `row.names` is a logical indicating if row names should be included and `quote` is another logical indicating if fields should be quoted. 

## Extra: csv types

There are two main types of csv files. In countries using the point for decimals, the delimiter is a comma. In countries using the comma for decimals, the delimiter is a semicolon. As a result, we have two functions for read and write csvs. 

- `read.csv` and `write.csv` reads and writes csv files from countries using decimal as point.

- `read.csv2` and `write.csv2` reads and writes csv files from countries using comma as point.

csv files are usually an intermediate file between Excel and R. Therefore, it is important to know your (or your colleagues) Excel configuration. It is very common that you experience problems when reading csv files in Excel created with R (e.g. small numbers are converted to very big numbers). If this is the case, try to create the csv file using the other version (e.g. if you created the file with `write.csv` and you have problems, repeat the process with `write.csv2`). The same may happen when reading csv files in R created with Excel. Again, changing the function used to read the file may solve the issue. 

# More resources on data management

You can find more information about data management in the following links:

- [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html): dplyr is a package designed to perform data management. Here you have some tutorials of how to use the package. 
- [tidyr](https://cran.r-project.org/web/packages/tidyr/index.html): tidyr is an another package to data management that works in conjunction with dplyr. It is useful to make conversions between wide and long formats. 
- [Cookbook for R: Manipulating Data](http://www.cookbook-r.com/Manipulating_data/): in this book you will find how to perform some common operations step by step. 

